Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-12 main.tex
sec.VII.A

\newcommand*{\LK}{\mathsf{LK}}
\newcommand*{\FOLK}{\mathsf{LK1}}
\newcommand*{\LKm}{\mathsf{LK + mix}}
\newcommand*{\MLL}{\mathsf{MLL}}
\newcommand*{\FOMLL}{\mathsf{MLL1^\times}} 
\newcommand*{\FOMLLm}{\mathsf{MLL1+mix}}
\newcommand*{\FOKS}{\mathsf{KS1}}
\newcommand*{\FOMLS}{\mathsf{MLS1^\times}}

@@

\newcommand{\ax}{\mathsf{ax}}
\newcommand{\cut}{\mathsf{cut}}
\newcommand{\conj}{\mathsf{\wedge}}
\newcommand{\disj}{\mathsf{\vee}}
\newcommand{\univ}{\mathsf{\forall}}
\newcommand{\exist}{\mathsf{\exists}}
\newcommand{\mix}{\mathsf{mix}}
\newcommand{\ctr}{\mathsf{ctr}}
\newcommand{\wk}{\mathsf{wk}}

@@

@@

We define a congruence relation $\fequ$ on formulas by the following equations:
\begin{equation*}
  \begin{array}{cc}
    A \vlan B \fequ B \vlan A & A \vlor B \fequ B \vlor A \\
    (A \vlan B) \vlan C \fequ A \vlan (B \vlan C) &
    (A \vlor B) \vlor C \fequ A \vlor (B \vlor C) \\
    \forall x . \forall y . A \fequ \forall y . \forall x . A &
    \exists x . \exists y . A \fequ \exists y . \exists x . A \\
    \forall x . (A \vlor B) \fequ (\forall x . A) \vlor B &
    \exists x . (A \vlan B) \fequ (\exists x . A) \vlan B
  \end{array}
\end{equation*}
where $x \notin fv(B)$ in the last two equations.
Two formulas $A$ and $B$ are \bfit{equivalent} if $A \fequ B$. The following
theorem shows that the set of fographs can been seen as the quotient
$\FORM / \fequ$.
\begin{thm}
  Let $A,B$ be rectified formulas. Then
  $$
  A\fequ B \iff \graphof A =\graphof B
  $$
\end{thm}

@@

In this section we prove Theorems
\subsection{The Linear Fragments $\FOMLL$ and $\FOMLS$}

>>>>

\newcommand*{\LK}{\mathsf{LK}}
\newcommand*{\FOLK}{\mathsf{LK1}}
\newcommand*{\FOLKcut}{\FOLK\mathord+\cut}
\newcommand*{\LKm}{\mathsf{LK + mix}}
\newcommand*{\MLL}{\mathsf{MLL}}
\newcommand*{\FOMLL}{\mathsf{MLL1^X}} 
\newcommand*{\FOMLLcut}{\FOMLL\mathord+\cut}
\newcommand*{\FOMLLm}{\mathsf{MLL1+mix}}
\newcommand*{\FOKS}{\mathsf{KS1}}
\newcommand*{\FOMLS}{\mathsf{MLS1^X}}

@@

\newcommand{\rr}{\mathsf{r}}
\newcommand{\ax}{\mathsf{ax}}
\newcommand{\cut}{\mathsf{cut}}
\newcommand{\conj}{\mathsf{\wedge}}
\newcommand{\disj}{\mathsf{\vee}}
\newcommand{\univ}{\mathsf{\forall}}
\newcommand{\exist}{\mathsf{\exists}}
\newcommand{\mix}{\mathsf{mix}}
\newcommand{\ctr}{\mathsf{ctr}}
\newcommand{\wk}{\mathsf{wk}}

@@

In this paper we make also use of the cut elimination theorem. The \bfit{cut} rule is
\begin{equation}
  \vliinf{\cutr}{}{\sqn{\Gamma,\Delta}}{\sqn{\Gamma,A}}{\sqn{\dual A,\Delta}}
\end{equation}

\begin{thm}
  \label{thm:cutelim}
  If a sequent $\sqn\Gamma$ is provable in $\FOLKcut$ then it is also
  provable in $\FOLK$. Furthermore, if $\sqn\Gamma$ is provable in
  $\FOMLLcut$ then it is also provable in $\FOMLL$.
\end{thm}

As before, this is standard, see e.g.~\cite{TS:00} for a proof.

@@

We define a congruence relation on formulas by the following equations:
\begin{equation*}
  \begin{array}{cc}
    A \vlan B \fequ B \vlan A & A \vlor B \fequ B \vlor A \\
    (A \vlan B) \vlan C \fequ A \vlan (B \vlan C) &
    (A \vlor B) \vlor C \fequ A \vlor (B \vlor C) \\
    \forall x . \forall y . A \fequ \forall y . \forall x . A &
    \exists x . \exists y . A \fequ \exists y . \exists x . A \\
    \forall x . (A \vlor B) \fequ (\forall x . A) \vlor B &
    \exists x . (A \vlan B) \fequ (\exists x . A) \vlan B
  \end{array}
\end{equation*}
where $x \notin fv(B)$ in the last two equations.

\begin{thm}
  Let $A,B$ be rectified formulas. Then
  $$
  A\fequ B \iff \graphof A =\graphof B
  $$
\end{thm}

@@

In this section we prove Theorems~\ref{thm:KS1},
\ref{thm:decomposition}, and~\ref{thm:decompositionA}, mainly by
translating derivations to and from the sequent calculus, and by rule
permutation arguments.

\subsection{The Linear Fragments $\FOMLL$ and $\FOMLS$}

In this section we show the equivalence of $\FOMLL$ and $\FOMLS$.

\begin{lemma}\label{lem:MLL1->MLS1}
  If $\sqn\Gamma$ is rectified and provable in $\FOMLL$ then $\form\Gamma$ is provable in  $\FOMLS$. \todo{check: do we need rectified here?}
\end{lemma}

\begin{proof}
  This is a straightforward induction on the proof of $\sqn\Gamma$ in
  $\FOMLL$, making a case analysis on the bottommost rule instance. We
  show here only the case of $\vlinf{\forall}{}{\sqn{\Delta,\forall
      x.A}}{\sqn{\Delta,A}}$ (all other cases are simpler and have
  been shown before, e.g.~\cite{brunnler:phd}): By induction
  hypothesis, there is a proof of $\form\Delta\vlor A$ in $\FOMLS$. We
  can prefix every line in that proof by $\forall x$ and then compose
  the following derivation:
  \begin{equation*}
    \vlderivation{
      \vlin{\fequ}{}{\form\Delta\vlor\forall x.A}{
        \vlde{\FOMLS}{}{\forall x.\form\Delta\vlor A}{
          \vlin{\forall}{}{\forall x.\ttt}{
            \vlhy{\ttt}}}}}
  \end{equation*}
  where we can apply the $\fequ$-rule because $x$ is not free in $\Delta$.
\end{proof}

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}

\begin{proof}
  This is a straightforward exercise that we leave to the reader.
\end{proof}

\begin{lemma}\label{lem:context}
  Let $A,B$ be formulas, and let $\Sconhole$ be a (positive)
  context. If $\sqn{\cneg A,B}$ is provable in $\FOMLL$, then so is
  $\sqn{\cneg{\Scons A},\Scons B}$.
\end{lemma}

\begin{proof}
  Straightforward induction on  $\Sconhole$. (see e.g. \cite{gug:str:01})
\end{proof}

\begin{lemma}\label{lem:MLS1->MLL1}
  If a formula $C$ is provable in $\FOMLS$ then $\sqn C$ is provable in  $\FOMLL$. \todo{check: do we need rectified here?}
\end{lemma}

\begin{proof}
  We proceed by induction on the number of inference steps in the
  proof of $C$ in $\FOMLS$. Consider the bottommost rule instance
  $\vlinf{\rr}{}{\Scons B}{\Scons A}$. By induction hypothesis we have
  a $\FOMLL$ proof of $\sqn{\Scons A}$, and by
  Lemma~\ref{lem:context}, we have a $\FOMLL$ proof of
  $\sqn{\cneg{\Scons A},\Scons B}$. We can compose them via $\cutr$:
  \begin{equation*}
    \vliinf{\cutr}{}{\sqn{\Scons B}}{\sqn{\Scons A}}{\sqn{\cneg{\Scons
          A},\Scons B}}
  \end{equation*}
  and then apply Theorem~\ref{thm:cutelim}.
\end{proof}
